---
title: "Metabolomics — Build Phyloseq Object"
output: html_document
date: "2025-03-14"
author: "Vanja Milenkovic"
---

> **Run this script first.** All downstream scripts (01–07) depend on the
> objects and configuration variables created here. Open this file in RStudio
> and knit/source it before running any other script in the pipeline.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Required packages

```{r load-packages}
library(phyloseq)
library(tibble)
library(dplyr)
library(here)   # portable paths — install with: install.packages("here")
```

---

## ── USER CONFIGURATION ────────────────────────────────────────────────────

Edit this block to match your data. All downstream scripts read these
variables — you should **not** need to change column names anywhere else.

```{r user-config}
# ── Input file paths ──────────────────────────────────────────────────────────
# Place your files in the project folder, or provide paths relative to it.
file_abundance <- here("glasshouse_stem_STATS.csv")   # feature abundance table
file_taxonomy  <- here("glasshouse_stem_TAX.csv")     # chemical classification
file_metadata  <- here("glasshouse_stem_META.csv")    # sample metadata

# ── Identifier columns ────────────────────────────────────────────────────────
col_feature_id     <- "row.ID"              # row identifier in the abundance table
col_feature_id_tax <- "mappingFeatureId"    # row identifier in the taxonomy table
col_sample_id      <- "filename"            # sample identifier in the metadata table

# ── Taxonomy columns (used in scripts 04, 05, 06) ────────────────────────────
col_chem_name   <- "ChemVistaName"                   # preferred chemical name
col_tax_fallback <- "ClassyFire.most.specific.class" # fallback name if col_chem_name is empty
col_tax_level   <- "ClassyFire.class"                # compound class level for composition plots

# ── Metadata variables for plots ─────────────────────────────────────────────
col_color <- "timepoint"        # primary grouping variable  → colour in plots
col_shape <- "group_treatment"  # secondary grouping variable → shape in plots

# ── Samples to exclude from some analyses ────────────────────────────────────
# Set col_exclude_val to NULL to include all samples (no exclusion).
col_exclude_var <- "timepoint"    # metadata column that contains the level to exclude
col_exclude_val <- "T3 healthy"   # value to exclude (e.g. a reference/baseline group)
#   col_exclude_val <- NULL       # ← uncomment this line to disable exclusion

# ── Variables for PERMANOVA (script 03) ───────────────────────────────────────
# List every metadata column you want to test as a main effect.
col_stat_vars <- c("timepoint", "infection", "treatment")

# Full model: all main effects + all interactions between col_stat_vars.
# Adjust if your design has fewer or different interaction terms.
formula_full <- "~ timepoint + infection + treatment +
                   timepoint:infection + timepoint:treatment + infection:treatment +
                   timepoint:infection:treatment"

# Best model: keep only the significant terms from the full model (edit after
# inspecting full_model results in script 03).
formula_best <- "~ timepoint + infection + treatment +
                   infection:treatment + timepoint:treatment"

# Formula for the constrained RDA in script 02 (same structure as formula_best
# unless you want to visualise a different model).
formula_rda <- "~ timepoint + treatment + infection + treatment:infection"

# ── limma group factor (script 07) ───────────────────────────────────────────
# Original group labels (must match values in the col_shape column of your metadata).
limma_levels <- c("Control AOD", "Drought AOD", "Control water", "Drought water")
# Safe labels (no spaces) used internally by limma's model matrix.
limma_labels <- c("Control_AOD",  "Drought_AOD",  "Control_water", "Drought_water")
# Contrast to test (use the safe label names defined above).
limma_contrast <- "Drought_AOD - Drought_water"
```

---

## Input files

Three CSV files are required. The table below shows the expected structure —
update `user-config` above to match your own file names and column names.

| File | Contents |
|------|----------|
| Abundance (`file_abundance`) | Feature intensity table from MZmine — features in rows, samples in columns. Must have a feature ID column (`col_feature_id`). |
| Taxonomy (`file_taxonomy`) | Chemical classification from SIRIUS/MolNetEnhancer/CANOPUS exported from Cytoscape — features in rows. Must have a feature ID column (`col_feature_id_tax`). |
| Metadata (`file_metadata`) | Sample metadata in GNPS format (without the `ATTRIBUTE_` prefix). Must have a sample ID column (`col_sample_id`) matching the sample column names in the abundance table. |

```{r load-data}
otu_mat    <- read.csv(file_abundance, stringsAsFactors = FALSE)
tax_mat    <- read.csv(file_taxonomy)
samples_df <- read.csv(file_metadata)
```

## Build phyloseq object

```{r build-phyloseq}
otu_mat    <- otu_mat    %>% tibble::column_to_rownames(col_feature_id)
tax_mat    <- tax_mat    %>% tibble::column_to_rownames(col_feature_id_tax)
samples_df <- samples_df %>% tibble::column_to_rownames(col_sample_id)

OTU     <- otu_table(as.matrix(otu_mat),    taxa_are_rows = TRUE)
TAX     <- tax_table(as.matrix(tax_mat))
samples <- sample_data(samples_df)

oakgs <- phyloseq(OTU, TAX, samples)
ntaxa(oakgs)   # number of metabolite features
```

## Create working objects for downstream scripts

```{r create-objects}
# Full dataset
pseq <- oakgs

# Dataset with the baseline/reference group excluded.
# This object is used in scripts where the reference group would confound
# treatment comparisons. Set col_exclude_val to NULL in the config above to
# keep all samples.
if (!is.null(col_exclude_val)) {
  pseqs <- subset_samples(oakgs, sample_data(oakgs)[[col_exclude_var]] != col_exclude_val)
  message("Created `pseqs`: ", col_exclude_val, " removed from ", col_exclude_var,
          " (", nsamples(pseqs), " of ", nsamples(oakgs), " samples retained)")
} else {
  pseqs <- oakgs
  message("No exclusion applied — `pseqs` is identical to `pseq`")
}

rm(otu_mat, samples, samples_df, tax_mat, OTU, TAX)
```
