---
title: "Principal Component Analysis (PCA)"
output: html_document
date: "2025-03-24"
author: "Vanja Milenkovic"
---

> **Prerequisite:** Run `00.phyloseq_object.Rmd` first. That script creates
> the `pseq` / `pseqs` objects and the configuration variables used here
> (`col_color`, `col_shape`, `col_exclude_var`, `col_exclude_val`).

PCA is an unconstrained ordination method that operates directly on
(transformed) feature variables without a distance matrix. A centred
log-ratio (CLR) transformation is applied before PCA to handle the
compositional nature of metabolomics data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r load-packages}
library(plyr)
library(dplyr)
library(phyloseq)
library(ggplot2)
library(microbiomeutilities)
library(microbiome)
library(microViz)
library(patchwork)
```

## Prepare datasets

CLR transformation is required for PCA. Two datasets are prepared:

- `ps_all` — all samples
- `ps_timepoints` — reference/baseline group excluded (controlled by
  `col_exclude_var` / `col_exclude_val` in script 00)

```{r prepare-data}
ps_all <- microbiome::transform(pseq, "clr")

if (!is.null(col_exclude_val)) {
  ps_timepoints <- ps_filter(pseq, .data[[col_exclude_var]] != col_exclude_val)
  ps_timepoints <- microbiome::transform(ps_timepoints, "clr")
} else {
  ps_timepoints <- ps_all
}
```

## PCA — all samples

```{r pca-all-samples}
theme_set(theme_bw())

pca_all <- ps_all %>%
  ord_calc() %>%
  ord_plot(color = col_color, shape = col_shape, size = 2) +
  scale_colour_brewer(palette = "Set1") +
  stat_ellipse(aes(colour = .data[[col_color]], group = .data[[col_color]]),
               size = 0.5) +
  labs(
    title = "Principal Component Analysis (PCA) — all samples",
    color = col_color,
    shape = col_shape
  ) +
  theme(
    axis.text.x     = element_text(size = 6, hjust = 1),
    axis.text.y     = element_text(size = 6),
    axis.title.x    = element_text(size = 8),
    axis.title.y    = element_text(size = 8),
    plot.title      = element_text(size = 10),
    legend.text     = element_text(size = 5),
    legend.title    = element_text(size = 7),
    legend.key.size = unit(0.3, "cm")
  )

pca_all

dir.create("plots", showWarnings = FALSE)
ggsave("plots/PCA_all_metabolites.png", pca_all,
       width = 6.3, height = 3.15, units = "in", dpi = 600, bg = "white")
```

## PCA — reference group excluded

Only produced when `col_exclude_val` is not NULL in script 00.

```{r pca-timepoints}
pca_t <- ps_timepoints %>%
  ord_calc() %>%
  ord_plot(color = col_color, shape = col_shape, size = 2) +
  scale_colour_brewer(palette = "Set1") +
  stat_ellipse(aes(colour = .data[[col_color]], group = .data[[col_color]])) +
  labs(
    title = paste0("PCA — ", col_exclude_val %||% "all", " excluded"),
    color = col_color,
    shape = col_shape
  ) +
  theme(legend.position = "none")

pca_t
```
