---
title: "Ordination — NMDS, PCoA, and RDA"
output: html_document
date: "2025-03-18"
author: "Vanja Milenkovic"
---

> **Prerequisite:** Run `00.phyloseq_object.Rmd` first. That script creates
> the `pseq` object and the configuration variables used here (`col_color`,
> `col_shape`, `formula_rda`).

Three ordination methods are used:

- **NMDS** (Non-metric Multidimensional Scaling) — distance-based, non-linear
- **PCoA** (Principal Coordinates Analysis) — distance-based, linear
- **RDA** (Redundancy Analysis) — constrained ordination combining PCA with
  regression; uses CLR-transformed data and the best-fit model from
  `03.permanova.Rmd` (defined as `formula_rda` in script 00)

Bray-Curtis dissimilarities are used for NMDS and PCoA. CLR transformation
is used for RDA.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r load-packages}
library(plyr)
library(dplyr)
library(phyloseq)
library(ggplot2)
library(microbiomeutilities)
library(microbiome)
library(microViz)
library(patchwork)
```

## NMDS and PCoA

Bray-Curtis dissimilarities require relative abundance (compositional)
data — CLR is not appropriate here.

```{r nmds-pcoa-compute}
ps_comp <- microbiome::transform(pseq, "compositional")

dist      <- "bray"
ord_meths <- c("NMDS", "PCoA")

plist <- llply(as.list(ord_meths), function(i, physeq, dist) {
  ordi <- ordinate(physeq, method = i, distance = dist)
  plot_ordination(physeq, ordi, "samples",
                  color = col_color, shape = col_shape)
}, ps_comp, dist)

names(plist) <- ord_meths
```

```{r nmds-pcoa-combine}
pdataframe <- ldply(plist, function(x) {
  df <- x$data[, 1:2]
  colnames(df) <- c("Axis_1", "Axis_2")
  cbind(df, x$data)
})
names(pdataframe)[1] <- "method"
```

```{r nmds-pcoa-plot}
p <- ggplot(pdataframe,
            aes(Axis_1, Axis_2,
                color = .data[[col_color]],
                shape = .data[[col_shape]],
                fill  = .data[[col_color]])) +
  geom_point(size = 2) +
  facet_wrap(~method, scales = "free") +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  scale_colour_brewer(type = "qual", palette = "Set1") +
  stat_ellipse(aes(colour = .data[[col_color]], group = .data[[col_color]]),
               size = 0.1) +
  labs(
    title  = "Ordination of Metabolite Composition (Bray-Curtis dissimilarities)",
    x      = "Axis 1",
    y      = "Axis 2",
    color  = col_color,
    fill   = col_color,
    shape  = col_shape
  ) +
  theme_bw() +
  theme(
    axis.text.x     = element_text(size = 6, angle = 45, hjust = 1),
    axis.text.y     = element_text(size = 6),
    axis.title.x    = element_text(size = 8),
    axis.title.y    = element_text(size = 8),
    plot.title      = element_text(size = 10),
    strip.text      = element_text(size = 6),
    legend.text     = element_text(size = 5),
    legend.title    = element_text(size = 7),
    legend.key.size = unit(0.3, "cm")
  )

p

dir.create("plots", showWarnings = FALSE)
ggsave("plots/ordination_NMDS_PCoA.png", p,
       width = 6.3, height = 3.15, units = "in", dpi = 600, bg = "white")
```

## RDA — constrained ordination

RDA uses CLR-transformed data. The model formula (`formula_rda`) is defined
in script 00 and should reflect your best-fit PERMANOVA model from script 03.

```{r rda-plot}
ps_rda <- microbiome::transform(ps_comp, "clr")

ord <- ordinate(ps_rda, method = "RDA",
                formula = as.formula(formula_rda))

p_rda <- plot_ordination(ps_rda, ord,
                         color = col_color, shape = col_shape) +
  scale_colour_brewer(type = "qual", palette = "Set1") +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  geom_point(size = 3) +
  stat_ellipse(aes(group = .data[[col_color]])) +
  ggtitle(paste("RDA:", formula_rda))

p_rda
```

## Individual NMDS and PCoA plots

```{r nmds-individual}
p_nmds <- plist[[1]] +
  scale_colour_brewer(type = "qual", palette = "Set1") +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  geom_point(size = 3) +
  stat_ellipse(aes(colour = .data[[col_color]], group = .data[[col_color]])) +
  ggtitle(names(plist)[1])

p_nmds
```

```{r pcoa-individual}
p_pcoa <- plist[[2]] +
  scale_colour_brewer(type = "qual", palette = "Set1") +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  geom_point(size = 3) +
  stat_ellipse(aes(colour = .data[[col_color]], group = .data[[col_color]])) +
  ggtitle(names(plist)[2])

p_pcoa
```
