---
title: "Bubble Plot — Pairwise PERMANOVA Results"
output: html_document
date: "2025-04-24"
author: "Vanja Milenkovic"
---

> **Prerequisite:** Run `03.permanova.Rmd` first to generate the pairwise
> PERMANOVA results. This script reads a pre-formatted summary CSV and
> visualises the results as a bubble plot.

Bubble plots show R² (bubble size) and p-value (bubble colour) for each
pairwise group comparison, split by experimental factor category.

**Required input file:** `permanova_final_bubble_data.csv`

This CSV must contain the following columns:

| Column | Description |
|--------|-------------|
| `group1` | First group in the pairwise comparison |
| `group2` | Second group in the pairwise comparison |
| `R2` | Effect size (R²) from pairwise PERMANOVA |
| `p_value` | p-value from pairwise PERMANOVA |
| `significance` | Significance label (e.g. `*`, `**`, or empty string) |
| `category` | Factor category label (e.g. `"Timepoint"`, `"Treatment"`, `"Interaction"`) |

The `category` values drive the panel split — edit `bubble_categories` below
to match the categories in your CSV.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ── USER CONFIGURATION ───────────────────────────────────────────────────────

```{r bubble-config}
# Path to the pre-formatted pairwise PERMANOVA results CSV
file_bubble <- here::here("permanova_final_bubble_data.csv")

# Category names to display as separate panels (must match values in the
# `category` column of file_bubble). Order controls the panel order.
bubble_categories <- c("Timepoint", "Treatment", "Interaction")
```

---

## Load packages

```{r load-packages}
library(ggplot2)
library(dplyr)
library(patchwork)
library(here)
```

## Load and prepare data

```{r load-data}
df <- read.csv(file_bubble)

df$category <- factor(df$category, levels = bubble_categories)
```

## Build bubble plots

```{r bubble-helper}
make_bubble <- function(data) {
  ggplot(data, aes(x = group1, y = group2, size = R2, fill = p_value)) +
    geom_point(shape = 21, color = "black", alpha = 0.7) +
    geom_text(
      data = subset(data, significance != ""),
      aes(label = significance),
      color = "white", size = 4, fontface = "bold"
    ) +
    scale_size_continuous(
      limits = c(0.02, 0.4), range = c(3, 15), name = "R²"
    ) +
    scale_fill_viridis_c(
      name      = "p-value",
      option    = "magma",
      direction = -1,         # darker = smaller p-value
      limits    = c(0, 0.06),
      oob       = scales::squish
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      axis.title.x = element_blank(),
      axis.title.y = element_blank()
    )
}
```

```{r build-panels}
panel_list <- lapply(seq_along(bubble_categories), function(i) {
  cat_label <- bubble_categories[i]
  make_bubble(filter(df, category == cat_label)) +
    ggtitle(paste0(LETTERS[i], ". ", cat_label))
})
```

## Combined figure

```{r combined-figure, fig.width=12, fig.height=10}
# Layout: first two panels side by side, remaining panels below
n <- length(panel_list)
if (n >= 2) {
  top_row   <- Reduce(`+`, panel_list[1:2])
  if (n > 2) {
    bottom_row <- Reduce(`+`, panel_list[3:n])
    combined   <- top_row / bottom_row +
      plot_layout(guides = "collect", heights = c(1, n - 2)) &
      theme(plot.title = element_text(size = 16, face = "bold"),
            legend.position = "right")
  } else {
    combined <- top_row +
      plot_layout(guides = "collect") &
      theme(plot.title = element_text(size = 16, face = "bold"),
            legend.position = "right")
  }
} else {
  combined <- panel_list[[1]]
}

print(combined)
```

## Save figure

```{r save-figure}
dir.create("plots", showWarnings = FALSE)
ggsave("plots/permanova_bubble_plot.png", combined,
       width = 12, height = 10, dpi = 600)
```
