---
title: "PERMANOVA — Metabolome Composition Statistics"
output: html_document
date: "2025-03-19"
author: "Vanja Milenkovic"
---

> **Prerequisite:** Run `00.phyloseq_object.Rmd` first. That script creates
> the `pseq` object and the configuration variables used here:
> `col_stat_vars`, `formula_full`, `formula_best`,
> `col_exclude_var`, `col_exclude_val`.

PERMANOVA (Permutational Multivariate Analysis of Variance) tests whether
defined grouping variables significantly explain variation in overall metabolome
composition, using Bray-Curtis dissimilarities on compositionally transformed
data.

Statistical tests are run twice: on the full dataset and on the dataset with
the reference group excluded (controlled by `col_exclude_var`/`col_exclude_val`
in script 00).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r load-packages}
library(phyloseq)
library(ggplot2)
library(vegan)
library(microbiome)
library(RColorBrewer)
library(dplyr)
library(pairwiseAdonis)
```

## Prepare datasets

```{r prepare-data}
ps0  <- microbiome::transform(pseq, "compositional")
meta <- meta(ps0)
dist_mat <- distance(ps0, method = "bray")

if (!is.null(col_exclude_val)) {
  ps.t     <- ps_filter(pseq, .data[[col_exclude_var]] != col_exclude_val)
  ps.t     <- microbiome::transform(ps.t, "compositional")
  meta.t   <- meta(ps.t)
  dist_mat.t <- distance(ps.t, method = "bray")
} else {
  ps.t     <- ps0
  meta.t   <- meta
  dist_mat.t <- dist_mat
}
```

## Single-factor PERMANOVA — full dataset

Each variable in `col_stat_vars` is tested independently.

```{r permanova-single-full}
for (v in col_stat_vars) {
  cat("\n──", v, "──\n")
  f <- as.formula(paste("dist_mat ~", v))
  print(adonis2(f, data = meta, permutations = 9999))
}
```

## Single-factor PERMANOVA — reference group excluded

```{r permanova-single-no-ref}
if (!is.null(col_exclude_val)) {
  for (v in col_stat_vars) {
    cat("\n──", v, "(", col_exclude_val, "excluded) ──\n")
    f <- as.formula(paste("dist_mat.t ~", v))
    print(adonis2(f, data = meta.t, permutations = 9999))
  }
}
```

## Model selection — full dataset

Start with the saturated model including all main effects and interactions
(`formula_full`, defined in script 00). Then remove non-significant terms
to arrive at the best-fit model (`formula_best`).

**After running `full_model`, update `formula_best` in script 00 to keep
only the significant terms, then re-run this script.**

```{r full-model}
full_model <- adonis2(
  as.formula(paste("dist_mat", formula_full)),
  data = meta, permutations = 9999, by = "term"
)
full_model
```

```{r best-model}
best_model <- adonis2(
  as.formula(paste("dist_mat", formula_best)),
  data = meta, permutations = 9999, by = "term"
)
best_model
```

## Model selection — reference group excluded

```{r full-model-no-ref}
if (!is.null(col_exclude_val)) {
  full_model_t <- adonis2(
    as.formula(paste("dist_mat.t", formula_full)),
    data = meta.t, permutations = 9999, by = "term"
  )
  full_model_t
}
```

```{r best-model-no-ref}
if (!is.null(col_exclude_val)) {
  best_model_t <- adonis2(
    as.formula(paste("dist_mat.t", formula_best)),
    data = meta.t, permutations = 9999, by = "term"
  )
  best_model_t
}
```

## Compare models — R² and adjusted R²

```{r compare-models}
if (!is.null(col_exclude_val)) {
  r2_values <- data.frame(
    Model = c("best_model (reference excluded)", "best_model (full dataset)"),
    R2    = c(best_model_t$R2[1], best_model$R2[1])
  )
  print(r2_values)

  n_t <- nrow(meta.t);  n <- nrow(meta)
  p_t <- length(best_model_t$R2) - 1
  p   <- length(best_model$R2)   - 1
  r2_t <- best_model_t$R2[1];  r2 <- best_model$R2[1]

  adj_r2_values <- data.frame(
    Model       = c("best_model (reference excluded)", "best_model (full dataset)"),
    Adjusted_R2 = c(1 - ((1 - r2_t) / (1 - p_t / n_t)),
                    1 - ((1 - r2)   / (1 - p   / n)))
  )
  print(adj_r2_values)
}
```

## Variance partitioning

How much variation does each predictor in `col_stat_vars` explain?

```{r variance-partitioning}
n_terms <- length(col_stat_vars)

if (!is.null(col_exclude_val)) {
  var_comparison <- data.frame(
    Term                  = col_stat_vars,
    R2_reference_excluded = best_model_t$R2[seq_len(n_terms)],
    R2_full_dataset       = best_model$R2[seq_len(n_terms)]
  )
} else {
  var_comparison <- data.frame(
    Term            = col_stat_vars,
    R2_full_dataset = best_model$R2[seq_len(n_terms)]
  )
}
print(var_comparison)
```

## Export PERMANOVA results

```{r export-results}
dir.create("stats", showWarnings = FALSE)
write.csv(best_model, "stats/permanova_best.csv", row.names = TRUE)
if (!is.null(col_exclude_val)) {
  write.csv(best_model_t,
            paste0("stats/permanova_best_no_", col_exclude_val, ".csv"),
            row.names = TRUE)
}
```

## Post-hoc pairwise PERMANOVA

Identify which specific group pairs drive the significant effects above.
All variables in `col_stat_vars` are tested.

```{r pairwise-permanova}
for (v in col_stat_vars) {
  cat("\n── Pairwise PERMANOVA:", v, "──\n")
  print(pairwise.adonis(dist_mat, factors = meta[[v]]))
}
```

```{r pairwise-interactions}
# All pairwise combinations of stat variables are tested as interaction factors
if (length(col_stat_vars) >= 2) {
  pairs <- combn(col_stat_vars, 2, simplify = FALSE)
  for (pair in pairs) {
    int_col <- paste0(pair[1], "_x_", pair[2])
    meta[[int_col]] <- interaction(meta[[pair[1]]], meta[[pair[2]]])
    cat("\n── Pairwise PERMANOVA: interaction", pair[1], "×", pair[2], "──\n")
    res_int <- pairwise.adonis(dist_mat, factors = meta[[int_col]])
    print(res_int)
    write.csv(res_int,
              paste0("stats/pairwise_", int_col, ".csv"),
              row.names = TRUE)
  }
}
```
