---
title: "Plot Selected Features"
output: html_document
date: "2025-03-18"
author: "Vanja Milenkovic"
---

> **Prerequisite:** Run `00.phyloseq_object.Rmd` first. That script creates
> the `pseq` object and configuration variables used here (`col_color`,
> `col_shape`, `col_chem_name`, `col_tax_fallback`).

This script visualises the abundance of a user-defined feature subset.
Features are identified by their row IDs from the abundance table (e.g. VIP
features from OPLS-DA models in SIMCA). Feature IDs are replaced with
chemical names from the taxonomy table where available.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ── USER CONFIGURATION ───────────────────────────────────────────────────────

```{r feature-config}
# Row IDs of the features you want to plot. Replace with your own feature IDs.
# These are the top 12 VIP features shared between two OPLS-DA models.
features_of_interest <- c("382",  "366",  "746",  "391",  "400",  "339",
                           "1556", "405",  "1557", "3235", "401",  "445")

# Optional plot title
plot_title <- "Abundance of selected VIP features"
```

---

## Load packages

```{r load-packages}
library(phyloseq)
library(microbiome)
library(microbiomeutilities)
library(ggplot2)
library(ggpubr)
library(tidyr)
library(dplyr)
library(scales)
```

## Subset phyloseq to selected features

```{r subset-features}
ps_sub <- prune_taxa(taxa_names(pseq) %in% features_of_interest, pseq)

cat(ntaxa(ps_sub), "of", length(features_of_interest),
    "requested features found in the phyloseq object.\n")
```

## Rename features using chemical classification

Replace numeric row IDs with chemical names from the taxonomy table.
Falls back to `col_tax_fallback` (defined in script 00) when the primary
name column (`col_chem_name`) is empty, and to "Unknown" if both are missing.

```{r rename-features}
tax_table(ps_sub) <- as.matrix(tax_table(ps_sub))

# Primary name
feature_names <- as.character(tax_table(ps_sub)[, col_chem_name])

# Fallback name
missing <- is.na(feature_names) | feature_names == ""
feature_names[missing] <- as.character(
  tax_table(ps_sub)[missing, col_tax_fallback]
)

# Final fallback
feature_names[is.na(feature_names) | feature_names == ""] <- "Unknown"

# Append row ID to guarantee unique names
taxa_names(ps_sub) <- paste0(feature_names, "_", taxa_names(ps_sub))
```

## Melt to long format

```{r melt-data}
df_sub <- psmelt(ps_sub)
```

## Bar plot — abundance of selected features

```{r feature-barplot, fig.width=8, fig.height=6}
# Order features by total abundance (most abundant first)
otu_order <- df_sub %>%
  group_by(OTU) %>%
  summarise(total_abundance = sum(Abundance), .groups = "drop") %>%
  arrange(desc(total_abundance)) %>%
  pull(OTU)

df_sub <- df_sub %>%
  mutate(OTU = factor(OTU, levels = otu_order))

p_features <- ggplot(df_sub,
                     aes(x    = .data[[col_color]],
                         y    = Abundance,
                         fill = .data[[col_shape]])) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_brewer(palette = "Set2") +
  facet_wrap(reformulate("OTU")) +
  theme_bw() +
  labs(
    title = plot_title,
    x     = col_color,
    y     = "Abundance",
    fill  = col_shape
  ) +
  scale_y_continuous(labels = label_log()) +
  theme(
    axis.text.x     = element_text(size = 6, angle = 45, hjust = 1),
    axis.text.y     = element_text(size = 6),
    axis.title.x    = element_text(size = 8),
    axis.title.y    = element_text(size = 8),
    plot.title      = element_text(size = 10),
    strip.text      = element_text(size = 5),
    legend.text     = element_text(size = 5),
    legend.title    = element_text(size = 7),
    legend.key.size = unit(0.3, "cm")
  )

print(p_features)

dir.create("plots", showWarnings = FALSE)
ggsave("plots/selected_features.png", p_features,
       width = 6.3, height = 5, units = "in", dpi = 600, bg = "white")
```
