---
title: "Heatmaps — Selected Features"
output: html_document
date: "2025-03-19"
author: "Vanja Milenkovic"
---

> **Prerequisite:** Run `00.phyloseq_object.Rmd` first. That script creates
> the `pseq` object and configuration variables used here (`col_color`,
> `col_shape`).

Heatmaps display relative abundance of a user-defined feature subset across
samples, with samples annotated by the metadata variables defined in script 00.
Features can be sorted and the top N most abundant are displayed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ── USER CONFIGURATION ───────────────────────────────────────────────────────

```{r heatmap-config}
# Row IDs of the features to include in the heatmap. Replace with your own.
# These are VIP features from the treatment OPLS-DA model (SIMCA).
features_heatmap <- c("15703", "15674", "18129", "1089",  "9293",  "42",
                       "17962", "5485",  "11908", "10359", "975",   "10208")

# Number of top features to display (by abundance)
top_n_features <- 12

# Annotation colours for the col_color variable (edit keys to match your levels)
colour_primary <- c(
  "T0"         = "#66C2A5",
  "T1"         = "#FC8D62",
  "T2"         = "#8DA0CB",
  "T3"         = "#E78AC3",
  "T3 healthy" = "#A6D854"
)

# Annotation colours for the col_shape variable (edit keys to match your levels)
colour_secondary <- c(
  "Control AOD"   = "#CAB2D6",
  "Control water" = "#6A3D9A",
  "Drought AOD"   = "#FB9A99",
  "Drought water" = "#E31A1C"
)
```

---

## Load packages

```{r load-packages}
library(microbiome)
library(knitr)
library(phyloseq)
library(tidyverse)
library(dplyr)
library(RColorBrewer)
library(microbiomeutilities)
library(pheatmap)
library(microViz)
```

## Subset and arrange data

Samples are sorted by `col_color` then `col_shape` (defined in script 00).
The last `ps_arrange()` call is the primary sort.

```{r subset-arrange}
ps0 <- prune_taxa(taxa_names(pseq) %in% features_heatmap, pseq)

ps0 <- ps_arrange(ps0, .data[[col_color]], .target = "sample_data")
ps0 <- ps_arrange(ps0, .data[[col_shape]], .target = "sample_data")

sample_data(ps0)
```

## Define colour palettes

```{r define-colours}
# Gradient palette for feature abundance
grad_ab     <- colorRampPalette(c("#faf3dd", "#f7d486", "#5e6472"))
grad_ab_pal <- grad_ab(12)

# Annotation track colours — keys must match the actual levels in your data
meta_colors        <- list(colour_primary, colour_secondary)
names(meta_colors) <- c(col_color, col_shape)
```

## Heatmap

```{r heatmap, fig.width=9.45, fig.height=5}
pmaptop <- plot_taxa_heatmap(
  ps0,
  subset.top        = top_n_features,
  VariableA         = c(col_color, col_shape),
  heatcolors        = grad_ab_pal,
  cluster_rows      = FALSE,
  transformation    = "compositional",
  cluster_cols      = FALSE,
  show_colnames     = FALSE,
  annotation_colors = meta_colors
)

pmaptop

dir.create("plots", showWarnings = FALSE)
ggsave("plots/selected_features_heatmap.png", pmaptop$plot,
       width = 9.45, height = 5, units = "in", dpi = 600, bg = "white")
```
