---
title: "Metabolite Class Composition Bar Plots"
output: html_document
date: "2025-04-09"
author: "Vanja Milenkovic"
---

> **Prerequisite:** Run `00.phyloseq_object.Rmd` first. That script creates
> the `pseq` and `pseqs` objects and the configuration variables used here
> (`col_color`, `col_shape`, `col_tax_level`).

Stacked bar plots show the relative abundance of metabolite compound classes
across samples. The compound class column is defined by `col_tax_level` in
script 00 (default: `ClassyFire.class`). Rare classes below the detection and
prevalence thresholds are merged into an "Other" category.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ── USER CONFIGURATION ───────────────────────────────────────────────────────

```{r composition-config}
# Minimum relative abundance to retain a class (default: 1%)
detection_threshold <- 1/100

# Minimum prevalence (fraction of samples) to retain a class (default: 10%)
prevalence_threshold <- 10/100

# Maximum number of classes to show with individual colours (rest = "Other")
n_taxa_plot <- 15
```

---

## Load packages

```{r load-packages}
library(microbiome)
library(dplyr)
library(phyloseq)
library(RColorBrewer)
library(tidyverse)
library(microViz)
```

## Prepare data

```{r prepare-data}
ps <- transform(pseq, "compositional")
ps <- aggregate_rare(ps,
                     level      = col_tax_level,
                     detection  = detection_threshold,
                     prevalence = prevalence_threshold)
```

## Composition bar plot — phyloseq default

```{r barplot-phyloseq, fig.width=8, fig.height=5}
n_classes  <- length(unique(tax_table(ps)[, col_tax_level]))
getPalette <- colorRampPalette(brewer.pal(11, "Paired"))

p <- plot_bar(ps, x = col_color, fill = col_tax_level) +
  geom_bar(aes(fill = .data[[col_tax_level]]),
           stat = "identity", position = "stack") +
  scale_fill_manual(values = getPalette(n_classes)) +
  facet_wrap(reformulate(col_shape), scales = "free_x") +
  theme_classic() +
  labs(fill = "Compound class",
       x    = col_color) +
  theme(axis.text.x = element_text(angle = -90, hjust = 0))

p
```

## Composition bar plot — microViz

```{r barplot-microviz-simple, fig.width=8, fig.height=4}
ps %>%
  comp_barplot(tax_level = col_tax_level) +
  coord_flip()
```

```{r barplot-microviz-custom, fig.width=10, fig.height=5}
ps %>%
  comp_barplot(
    tax_level           = col_tax_level,
    facet_by            = col_shape,
    label               = NULL,
    n_taxa              = n_taxa_plot,
    other_name          = "Other compounds",
    order_with_all_taxa = TRUE,
    tax_order           = prev,
    merge_other         = FALSE,
    bar_width           = 0.7,
    bar_outline_colour  = "grey5"
  ) +
  coord_flip()
```

## Merge and compare two datasets

Merge two phyloseq objects (e.g. two tissues or experiments) for direct
visual comparison. Replace `dataset1` and `dataset2` with your own objects
and update the `dataset` labels below.

```{r merge-datasets}
# Replace pseq / pseqs with your two datasets as needed
dataset1 <- pseq  %>% ps_mutate(dataset = "Dataset 1")
dataset2 <- pseqs %>% ps_mutate(dataset = "Dataset 2")

dataset1 <- dataset1 %>%
  tax_fix(unknowns = c("", " ", "unknown")) %>%
  transform("compositional") %>%
  aggregate_rare(level      = col_tax_level,
                 detection  = detection_threshold,
                 prevalence = prevalence_threshold)

dataset2 <- dataset2 %>%
  tax_fix(unknowns = c("", " ", "unknown")) %>%
  transform("compositional") %>%
  aggregate_rare(level      = col_tax_level,
                 detection  = detection_threshold,
                 prevalence = prevalence_threshold)

combined <- phyloseq::merge_phyloseq(
  dataset1 %>% tax_agg(col_tax_level) %>% ps_get(),
  dataset2 %>% tax_agg(col_tax_level) %>% ps_get()
)

combined
```

```{r merged-barplot-facet, fig.width=8, fig.height=5}
combined %>%
  comp_barplot(col_tax_level, facet_by = "dataset", n_taxa = n_taxa_plot) +
  coord_flip()
```

```{r merged-barplot-grouped, fig.width=8, fig.height=6}
plots <- combined %>%
  comp_barplot(col_tax_level, n_taxa = n_taxa_plot,
               group_by = "dataset", bar_outline_colour = NA)

patchwork::wrap_plots(plots, nrow = 2, guides = "collect") &
  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank())
```

```{r merged-barplot-treatment, fig.width=10, fig.height=6}
plots <- combined %>%
  comp_barplot(col_tax_level, n_taxa = n_taxa_plot,
               facet_by = col_shape, group_by = "dataset",
               bar_outline_colour = NA)

patchwork::wrap_plots(plots, nrow = 2, guides = "collect") &
  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank())
```
