---
title: "Differential Abundance — Log2 Fold Change"
output: html_document
date: "2025-04-26"
author: "Vanja Milenkovic"
---

> **Prerequisite:** Run `00.phyloseq_object.Rmd` first. That script creates
> the `oakgs` / `pseqs` objects and the configuration variables used here
> (`col_shape`, `col_color`, `col_exclude_var`, `col_exclude_val`,
> `limma_levels`, `limma_labels`, `limma_contrast`).

Two approaches to log2 fold-change analysis are provided:

1. **limma-based (recommended):** Linear modelling with empirical Bayes
   moderation. Requires a log-transformed intensity matrix and a metadata
   CSV as separate input files.
2. **Exploratory (phyloseq-based):** Simple mean log2FC between groups
   directly from a phyloseq object. No statistical testing — for quick visual
   screening only.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r load-packages}
library(limma)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(ggrepel)
library(microbiome)
library(here)
```

---

## Part 1 — limma differential abundance

### Input files

| File | Contents |
|------|----------|
| `*_STATS_log.csv` | Log-transformed feature intensity matrix — features in rows, samples in columns. |
| `*_META_log.csv`  | Sample metadata with a grouping column matching `col_shape` from script 00. |

```{r load-limma-data}
# Edit these paths to point to your own files
expr_M <- read.csv(here("glasshouse_stem_STATS_log.csv"), row.names = 1)
meta   <- read.csv(here("glasshouse_stem_META_log.csv"))
```

### Build design matrix

Group labels are renamed to remove spaces (limma requirement). The levels
and labels are defined in script 00 (`limma_levels`, `limma_labels`).

```{r build-design}
meta[[col_shape]] <- factor(meta[[col_shape]],
                            levels = limma_levels,
                            labels = limma_labels)

design           <- model.matrix(~0 + meta[[col_shape]])
colnames(design) <- limma_labels
```

### Fit model and define contrast

The contrast to test is defined as `limma_contrast` in script 00.
Edit that variable to change the comparison.

```{r fit-model}
fit  <- lmFit(expr_M, design)
cont <- makeContrasts(contrasts = limma_contrast, levels = design)
fit2 <- contrasts.fit(fit, cont)
fit2 <- eBayes(fit2)

res  <- topTable(fit2, number = Inf)
```

### Classify features as up / down / NS

```{r classify-features}
res <- res %>%
  mutate(direction = case_when(
    adj.P.Val <= 0.05 & logFC >=  1 ~ "Up",
    adj.P.Val <= 0.05 & logFC <= -1 ~ "Down",
    TRUE                             ~ "NS"
  ))

cat("Contrast tested:           ", limma_contrast, "\n")
cat("Up-regulated metabolites:  ", sum(res$direction == "Up"),   "\n")
cat("Down-regulated metabolites:", sum(res$direction == "Down"), "\n")
```

### Inspect top features

```{r inspect-top-features}
up_feats   <- res %>% filter(direction == "Up")   %>% arrange(desc(logFC))
down_feats <- res %>% filter(direction == "Down") %>% arrange(logFC)

head(up_feats,   10)
head(down_feats, 10)
```

### Volcano plot

```{r volcano-plot, fig.width=8, fig.height=6}
res2 <- res %>% rownames_to_column(var = "Metabolite")

sig      <- res2 %>% filter(adj.P.Val < 0.05, abs(logFC) >= 1)
top_up   <- sig  %>% filter(logFC > 0) %>% arrange(desc(logFC)) %>% slice_head(n = 10)
top_down <- sig  %>% filter(logFC < 0) %>% arrange(logFC)       %>% slice_head(n = 10)
to_label <- bind_rows(top_up, top_down)

ggplot(res2, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(color = direction), alpha = 0.6) +
  scale_color_manual(values = c(Up = "red", Down = "blue", NS = "grey")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_text_repel(
    data          = to_label,
    aes(label     = Metabolite),
    size          = 3,
    box.padding   = 0.3,
    point.padding = 0.2,
    max.overlaps  = Inf
  ) +
  labs(
    x     = expression(log[2]*"(Fold-Change)"),
    y     = expression(-log[10]*"(FDR)"),
    color = NULL,
    title = paste("Volcano Plot —", limma_contrast,
                  "\n(Top 10 Up & Down labelled)")
  ) +
  theme_minimal()
```

---

## Part 2 — Exploratory log2FC from phyloseq (no statistical testing)

Computes simple mean log2FC between two groups directly from a phyloseq
object. Useful for a quick visual screen; not appropriate for reporting.

```{r exploratory-config}
# Define the two groups to compare (use the safe label names from limma_labels)
group_numerator   <- limma_labels[2]   # e.g. "Drought_AOD"
group_denominator <- limma_labels[1]   # e.g. "Control_AOD"
```

```{r exploratory-prepare}
# Exclude baseline timepoint if configured in script 00
if (!is.null(col_exclude_val)) {
  ps_explore <- subset_samples(
    oakgs,
    sample_data(oakgs)[[col_exclude_var]] != col_exclude_val
  )
} else {
  ps_explore <- oakgs
}

ps.rel <- microbiome::transform(ps_explore, "compositional")
df     <- psmelt(ps.rel)

df[[col_shape]] <- factor(df[[col_shape]],
                           levels = limma_levels,
                           labels = limma_labels)

mean.df <- df %>%
  group_by(OTU, .data[[col_shape]]) %>%
  summarise(mu = mean(Abundance), .groups = "drop") %>%
  pivot_wider(names_from  = col_shape,
              values_from = mu,
              values_fill = 0)
```

```{r exploratory-log2fc}
mean.df <- mean.df %>%
  mutate(
    log2FC = log2((.data[[group_numerator]]   + 1e-6) /
                  (.data[[group_denominator]] + 1e-6)),
    direction = case_when(
      log2FC >  1 ~ paste("Up in", group_numerator),
      log2FC < -1 ~ paste("Up in", group_denominator),
      TRUE        ~ "No change"
    )
  )

cat("Comparison:", group_numerator, "vs", group_denominator, "\n")
print(table(mean.df$direction))
```
